\sys: Controllable Program Regeneration

### 2025-07-28 docs/plos/draft.tex
Make codeblocks table less space
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

@@

@@

\begin{abstract}
  Software evolution tasks---like security hardening, porting critical components to safer 
  languages, or refactoring opaque legacy code---are time-consuming and brittle.
  Large language models (LLMs) have shown promise in automating such tasks,
  but their outputs are shaped by surface cues, lack
  semantic guarantees, and their usage often boil down to brittle prompting strategies.
This paper presents \sys, a system for controllable program regeneration that
  combines declarative property constraints with LLM-guided synthesis. Unlike
  prior approaches, \sys treats the input program as an untrusted artifact and
  explicitly projects it into a minimal set of verifiable properties, such as
  input/output behavior, execution traces, and API signatures. These properties
  serve as constraints during synthesis and are checked post hoc to ensure
  correctness.
\sys is evaluated on a diverse set of program transformation tasks
including security hardening, language translation, idiomatic rewriting, de-obfuscation,
and modular refactoring showing strong results, even on tasks where state-of-the-art LLMs fail.
\end{abstract}

@@

\heading{A legacy input handling function}
Zooming into the \ttt{cmd.c} file of the game codebase, 
which handles parsing and executing user commands provided by an in-game console.
This functionality is used by developers to test and debug the game, but also during regular gameplay
by game administrators or players where they can customize the game by spawning objects, changing game settings, \etc.
Such code is often notoriously difficult to write,
and there have been many instances of buffer overflow vulnerabilities being introduced because of it~\cite{CVE-2006-3400, CVE-2006-3401, CVE-2007-5248, CVE-2019-1010043}.
\begin{listing}
\begin{minted}[]{js}
#define MAX_C 1024
char *Cmd_Args(int cmd_argc, char **cmd_argv) {
static char cmd_args[MAX_C];
int i; cmd_args[0] = 0;
for ( i = 1 ; i < cmd_argc ; i++ ) {
 strcat( cmd_args, cmd_argv[i] );
 if ( i != cmd_argc-1 ) {
   strcat( cmd_args, " " );
 } }
	return cmd_args; }
\end{minted}
\caption{A string formatting function that concatenates command-line arguments into a single string.}
\end{listing}

\begin{wrapfigure}[6]{r}{.7\columnwidth}
\vspace{-10pt}
\hspace{-10pt}
\begin{minted}{prolog}
fsignt(p_, S) :- fsignt(p, S).
graph(p_, I, O) :- graph(p, I, O),
                   crash(p, I).
:- crash(p_, I), graph(p_, I, _).
\end{minted}
\end{wrapfigure}
To the right is a program written in the \sys DSL to express this intent.
This query requires the regenerated program \texttt{P\_} to exhibit the same
I/O behavior on representative inputs that would cause a buffer overflow in the original program $P$.
but not on the regenerated program $P^\prime$.
\sys first arrives the set of properties to extract from the
original program, given the query:
$\{\ttt{input(P, I)}, \ttt{graph(P, I, O)}, \ttt{signature(P, S)}\}$, where
\texttt{input(P, I)} means that \sys will use extract a set of inputs from the original program
by prompting an LLM with the program's source code and asking it to generate a set of inputs that
would cause a buffer overflow in the original program;
\texttt{graph(P, I, O)} extracts the client-observable I/O behavior of the program
by executing the original program on the generated inputs and recording the outputs;
and \texttt{signature(P, S)} extracts each function's signature using a language-aware parser.

After 30.4 seconds and one attempt, \sys synthesizes the following program:
\begin{minted}[]{js}
#define MAX_C 1024
char *Cmd_Args(int cmd_argc, char **cmd_argv) {
static char cmd_args[MAX_C];
int i; size_t remaining = MAX_C - 1;
cmd_args[0] = 0;
for (i = 1; i < cmd_argc; i++) {
  strncat(cmd_args, cmd_argv[i], remaining);
  remaining = MAX_C - strlen(cmd_args) - 1;
  if (i != cmd_argc - 1 && remaining > 0) {
    strncat(cmd_args, " ", remaining);
    remaining = MAX_C - strlen(cmd_args) - 1;
  } }
return cmd_args; }
\end{minted}

@@

\begin{wrapfigure}[7]{r}{.5\columnwidth}
% \vspace{-10pt}
\begin{minted}{prolog}
language(p_, javascript).
graph(p_, Ijs, Ojs) :-
     graph(p, Ic, Oc),
llm_translate(Ic, Ijs),
llm_translate(Oc, Ijs).
\end{minted}
\end{wrapfigure}
The query specifies that the regenerated program \texttt{P\_} must be
implemented in JavaScript and must preserve observable behavior over a
representative set of inputs and outputs.
To bridge differences in
representation between C and JavaScript, the query uses \texttt{llm\_translate}
to convert each input and output into equivalent forms suitable for the target language.
\sys first determines the minimal set of properties to extract from the
original function: $\{\texttt{graph(P, I, O)}\}$.
It then invokes \texttt{llm\_translate} to produce
corresponding JavaScript representations of each example input and output, such
that the translated function will reproduce identical string formatting
behavior.
After 29.4 seconds, having generated 20 translated input-output pairs, \sys synthesizes the following JavaScript function:

@@

\heading{An unidiomatic C function in a physics library}
The fast inverse square root routine, originally developed for the Quake III
engine~\cite{fast_inv_sqrt}, is included in the game's physics and vector math
library as an optimization for computing normalized vectors.
The implementation
exploits type punning to manipulate IEEE floats at the bit level—a technique
that relies on undefined behavior and reduces portability:
\begin{listing}
\begin{minted}{c}
float Q_rsqrt(float number) {
 long i; float x2, y;
 const float threehalfs = 1.5F;
 x2 = number * 0.5F; y  = number;
 i  = * ( long * ) &y; // evil bit level hack
 i  = 0x5f3759df - ( i >> 1 );
 y  = * ( float * ) &i;
 y  = y * ( threehalfs - ( x2 * y * y ) );
 return y; }
\end{minted}
\caption{An unidiomatic C implementation of the fast inverse square root function popularized by the game Quake III~\cite{fast_inv_sqrt}.}
\end{listing}
The query to the right uses lines of code as a coarse metric of
idiomaticity. \sys regenerates the program, minimizing this metric while
preserving observable behavior over a representative set of inputs which will be generated by an LLM instance.
\begin{wrapfigure}[4]{r}{.7\columnwidth}
  \begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
#min N: len(p_, N).
  \end{minted}
\end{wrapfigure}
\sys extracts the set of properties \texttt{signature(F)} and \texttt{graph(F)}
from the original function, where \texttt{signature(F)} is the function's
signature (including its name, return type, and arguments) and
\texttt{graph(F)} captures the I/O behavior on a generated set of
floating-point inputs.
In this case, where the goal is to optimize an aspect of the program, \sys
re-prompts the model to regenerate the program until it can make no further
improvements or observes a negligible difference after two consecutive
iterations.

@@

\begin{wrapfigure}[7]{r}{.68\columnwidth}
  \begin{minted}{prolog}
db('game.db').func(f1).func(f2).
sql_trc(p_,I,T) :- sql_trc(p,I,T),
                   graph(p,I,_).
graph(call(f2,call(f1),I),I,O):-
              graph(p,I,O).
  \end{minted}
\end{wrapfigure}

@@

\section{System Overview}
\label{sec:design}

@@

\begin{table*}[ht]
\centering
\caption{\textbf{Benchmark Overview}.}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r l l}
\toprule
Benchmark & N & Purpose & Sources \\
\midrule
Rosetta Code & 16 & Cross-language translation & \cite{rosettacode} \\
Short Utilities & 13 & Parity preservation & \cite{regbench2025} \\
SSCA & 3 & Malicious behavior removal & \cite{ohm2020backstabber,ev:eurosec:2022} \\
IOCCC & 2 & Deobfuscation & \cite{ioccc} \\
Microbenchmarks & 3 & Idiomatic rewriting, modularization & This paper \\
\bottomrule
\end{tabular*}
\label{tab:benchmarks}
\end{table*}

>>>>

\setminted{
    fontsize=\small
  }

@@

\setlength{\intextsep}{0pt} % vertical space above/below
\setlength{\columnsep}{5pt} % horizontal space between text and figure

@@

\begin{abstract}
  Software evolution tasks---like security hardening, porting critical components to safer 
  languages, or refactoring opaque legacy code---are brittle and time-consuming.
  Large language models (LLMs) have shown promise in automating such tasks,
  but their outputs are shaped by surface cues, lack
  semantic guarantees, and their usage often boil down to brittle prompting strategies.
This paper presents \sys, a system for controllable program regeneration that
  combines declarative property constraints with LLM-guided synthesis. Unlike
  prior approaches, \sys treats the input program as an untrusted artifact and
  explicitly projects it into a minimal set of verifiable properties, such as
  input/output behavior, execution traces, and API signatures. These properties
  serve as constraints during synthesis and are checked post hoc to ensure
  correctness.
\sys is evaluated on a diverse set of program transformation tasks
including security hardening, language translation, idiomatic rewriting, de-obfuscation,
and modular refactoring showing strong results, even on tasks where state-of-the-art LLMs fail.
\end{abstract}

% Since this is a position paper (and not a tool paper), I would spend most of the paper selling the Idea, and less on the system and language. I think the examples does this well (but the language still feels a little too magical).
% The second section should probably focus on problem (program to program transformations) and the gab (related work), explaining why current approaches fall short.
% The third section should introduce our idea for a better solution, with examples. I think the core of the idea is building a knowledge base of the programs which we want to transform between, and then use (out of many) an LLM as an imperfect bidirectional analysis.
% I think the idea of bidirectional analysis is somewhat new, and handeling imperfect information in a logical system is not easy. I think we need to address why we think this is possible.
% The fourth section could contain a plan for future work: implementation and evaluation.
% The fifth section is a discussion and conclusion.

@@

\heading{A legacy input handling function}
Zooming into the \ttt{cmd.c} file of the game codebase, 
which handles parsing and executing user commands provided by an in-game console.
This functionality is used by developers to test and debug the game, but also during regular gameplay
by game administrators or players where they can customize the game by spawning objects, changing game settings, \etc.
Such code is often notoriously difficult to write,
and there have been many instances of buffer overflow vulnerabilities being introduced because of it~\cite{CVE-2006-3400, CVE-2006-3401, CVE-2007-5248, CVE-2019-1010043}.
\begin{listing}
\begin{minted}[]{js}
#define MAX_C 1024
char *Cmd_Args(int argc, char **argv) {
 static char args[MAX_C];
 for (int i = 1; i < argc; i++) {
  strcat(args, argv[i]);
  if (i < argc - 1) strcat(args, " "); }
 return args; }
\end{minted}
\caption{A string formatting function that concatenates command-line arguments into a single string.}
\end{listing}

\begin{wrapfigure}[4]{r}{.64\columnwidth}
\begin{minted}{prolog}
fsignt(p_, S) :- fsignt(p, S).
graph(p_, I, O) :- graph(p, I, O),
                   crash(p, I).
:- crash(p_, I), graph(p_, I, _).
\end{minted}
\end{wrapfigure}
To the right is a program written in the \sys DSL to express this intent.
This query requires the regenerated program \texttt{P\_} to exhibit the same
I/O behavior on representative inputs that would cause a buffer overflow in the original program $P$.
but not on the regenerated program $P^\prime$.
\sys first arrives the set of properties to extract from the
original program, given the query:
$\{\ttt{input(P, I)}, \ttt{graph(P, I, O)}, \ttt{signature(P, S)}\}$, where
\texttt{input(P, I)} means that \sys will use extract a set of inputs from the original program
by prompting an LLM with the program's source code and asking it to generate a set of inputs that
would cause a buffer overflow in the original program;
\texttt{graph(P, I, O)} extracts the client-observable I/O behavior of the program
by executing the original program on the generated inputs and recording the outputs;
and \texttt{signature(P, S)} extracts each function's signature using a language-aware parser.

After 30.4 seconds and one attempt, \sys synthesizes the following program:
\begin{minted}[]{js}
#define MAX_C 1024
char *Args(int argc, char **argv) {
static char args[MAX_C];
int i; size_t remaining = MAX_C - 1;
args[0] = 0;
for (i = 1; i < argc; i++) {
  strncat(args, argv[i], remaining);
  remaining = MAX_C - strlen(args) - 1;
  if (i != argc - 1 && remaining > 0) {
    strncat(args, " ", remaining);
    remaining = MAX_C - strlen(args) - 1;
  } }
  return args; }
\end{minted}

@@

\begin{wrapfigure}[5]{r}{.46\columnwidth}
\begin{minted}{prolog}
language(p_, javascript).
graph(p_, Ijs, Ojs) :-
     graph(p, Ic, Oc),
llm_translate(Ic, Ijs),
llm_translate(Oc, Ijs).
\end{minted}
\end{wrapfigure}
The query specifies that the regenerated program \texttt{P\_} must be
implemented in JavaScript and must preserve observable behavior over a
representative set of inputs and outputs.
To bridge differences in
representation between C and JavaScript, the query uses \texttt{llm\_translate}
to convert each input and output into equivalent forms suitable for the target language.
\sys first determines the minimal set of properties to extract from the
original function: $\{\texttt{graph(P, I, O)}\}$.
It then invokes \texttt{llm\_translate} to produce
corresponding JavaScript representations of each example input and output, such
that the translated function will reproduce identical string formatting
behavior.
After 29.4 seconds, having generated 20 translated input-output pairs, \sys synthesizes the following JavaScript function:

@@

\heading{An unidiomatic C function in a physics library}
The fast inverse square root routine, originally developed for the Quake III
engine~\cite{fast_inv_sqrt}, is included in the game's physics and vector math
library as an optimization for computing normalized vectors.
The implementation
exploits type punning to manipulate IEEE floats at the bit level—a technique
that relies on undefined behavior and reduces portability:
\begin{listing}
\begin{minted}{c}
float Q_rsqrt(float number) {
 long i; float x2, y;
 const float threehalfs = 1.5F;
 x2 = number * 0.5F; y  = number;
 i  = * ( long * ) &y; // evil bit level hack
 i  = 0x5f3759df - ( i >> 1 );
 y  = * ( float * ) &i;
 y  = y * ( threehalfs - ( x2 * y * y ) );
 return y; }
\end{minted}
\caption{An unidiomatic C implementation of the fast inverse square root function popularized by the game Quake III~\cite{fast_inv_sqrt}.}
\end{listing}
The query to the right uses lines of code as a coarse metric of
idiomaticity. \sys regenerates the program, minimizing this metric while
preserving observable behavior over a representative set of inputs which will be generated by an LLM instance.
\begin{wrapfigure}[2]{r}{.7\columnwidth}
  \begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
#min N: len(p_, N).
  \end{minted}
\end{wrapfigure}
\sys extracts the set of properties \texttt{signature(F)} and \texttt{graph(F)}
from the original function, where \texttt{signature(F)} is the function's
signature (including its name, return type, and arguments) and
\texttt{graph(F)} captures the I/O behavior on a generated set of
floating-point inputs.
In this case, where the goal is to optimize an aspect of the program, \sys
re-prompts the model to regenerate the program until it can make no further
improvements or observes a negligible difference after two consecutive
iterations.

@@

\begin{wrapfigure}[5]{r}{.68\columnwidth}
  \begin{minted}{prolog}
db('game.db').func(f1).func(f2).
sql_trc(p_,I,T) :- sql_trc(p,I,T),
                   graph(p,I,_).
graph(call(f2,call(f1),I),I,O):-
              graph(p,I,O).
  \end{minted}
\end{wrapfigure}

@@

\section{System Sketch}
\label{sec:design}

@@

\begin{table}[ht]
\centering
\caption{\textbf{Benchmark Overview}.}
\begin{tabular}{@{\extracolsep{\fill}} l r l l}
\toprule
Benchmark & N & Purpose & Sources \\
\midrule
Rosetta Code & 16 & Cross-language translation & \cite{rosettacode} \\
Short Utilities & 13 & Parity preservation & \cite{regbench2025} \\
SSCA & 3 & Malicious behavior removal & \cite{ohm2020backstabber,ev:eurosec:2022} \\
IOCCC & 2 & Deobfuscation & \cite{ioccc} \\
% Microbenchmarks & 3 & Idiomatic rewriting, modularization & This paper \\
\bottomrule
\end{tabular}
\label{tab:benchmarks}
\end{table}
